---
format: html
---

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
library(here)
library(gridExtra)

# source bias function
source(here("occAssess_spatial_env_bias_function.R"))
```

# Run Bias assessments

This chunk identifies the regions and assessments missing (for spatial and environmental bias; and at family or species) and runs the analysis to update their results

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
# urls to be downloaded from Dropbox
occ_data = tibble(dataset = c("mdg_invaded_clean",
                              "jpn_invaded_clean",
                              "irl_Invaded_clean",
                              "aus_glonaf",
                              "gbr_invaded_clean",
                              "zaf_invaded_clean",
                              "nzl_glonaf_invaded_clean",
                              "usa_invaded_clean"), 
                  
                  drop_url = c("https://www.dropbox.com/scl/fi/3hx8gbgd7858cpubyo4f1/MDG-invaded-clean.rds?rlkey=98cs05vrn2s55igxuuxhj5qdt&dl=1",
                               "https://www.dropbox.com/scl/fi/qezmtd6oj1jf1lsvtkmux/JPN-invaded-clean.rds?rlkey=rs3d5mgwys2s1731if509ef1v&dl=1",
                               "https://www.dropbox.com/scl/fi/gw9dpvdifdgp3sksinngo/IRL-invaded-clean.rds?rlkey=boyzevieh6vc8vi3ucwwwjo0c&dl=1",
                               "https://www.dropbox.com/scl/fi/kizxr5oy7oe5d54kyw76p/AUS-Glonaf-invaded-clean.rds?rlkey=3ukbga0g9q7hjgowshl018n2s&dl=1",
                               "https://www.dropbox.com/scl/fi/n0ridwusbnqk2lnrm078i/GBR-invaded-clean-v1.rds?rlkey=tejxsv5l3tge1552ld76fajrr&dl=1",
                               "https://www.dropbox.com/scl/fi/5a0a1pepumq1t1ujwo7k8/ZAF-invaded-clean.rds?rlkey=ltdovycts12kvvyjt1smc1arg&dl=1",
                               "https://www.dropbox.com/scl/fi/ux9levicn4cx3pz43v9i0/NZL-Glonaf-invaded-clean.csv?rlkey=9223dy2cm3atbg0dgbp8s81y3&dl=1",
                               "https://www.dropbox.com/scl/fi/rxrutz13ogzpcu8h28ycu/USA-invaded-clean.rds?rlkey=ptf649kbc44nflu0nsv3sdvnh&dl=1"
                               )) |>
  na.omit()


# generate bias output files
tables = occ_data$dataset[6] 
for (db_table in tables) {
  Bias_assessment_function(db_table = db_table, 
                           periods_length = 10)
}
```

# Occurrence Periods

Occurrence records were divided into 10 year periods to facilitate assessment of *temporal* variations in the levels of biases over time.

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
## AUS GLONAF
directory_files_aus_glonaf = list.files(here("13.  bias assessment results", "aus_glonaf_invaded_clean"))
file_link_aus_glonaf = here("13.  bias assessment results", "aus_glonaf_invaded_clean", directory_files_aus_glonaf[grepl("periods_output", directory_files_aus_glonaf)])
periods_aus_glonaf = read.csv(file = file_link_aus_glonaf)

periods_range_aus_glonaf = data.frame(Period = 1:ncol(periods_aus_glonaf))
for (i in 1:ncol(periods_aus_glonaf)) {
  periods_range_aus_glonaf$range_years[i] = paste(min(periods_aus_glonaf[, i], na.rm = T), 
                                  max(periods_aus_glonaf[ , i], na.rm = T), sep = "-")
}
periods_range_aus_glonaf$range_years = as.factor(periods_range_aus_glonaf$range_years)

## GBR
directory_files_gbr = list.files(here("13.  bias assessment results", "gbr_invaded_clean"))
file_link_gbr = here("13.  bias assessment results", "gbr_invaded_clean", directory_files_gbr[grepl("periods_output", directory_files_gbr)])
periods_gbr = read.csv(file = file_link_gbr)

periods_range_gbr = data.frame(Period = 1:ncol(periods_gbr))
for (i in 1:ncol(periods_gbr)) {
  periods_range_gbr$range_years[i] = paste(min(periods_gbr[, i], na.rm = T), 
                                  max(periods_gbr[ , i], na.rm = T), sep = "-")
}
periods_range_gbr$range_years = as.factor(periods_range_gbr$range_years)

## IRL
directory_files_irl = list.files(here("13.  bias assessment results", "irl_invaded_clean"))
file_link_irl = here("13.  bias assessment results", "irl_invaded_clean", directory_files_irl[grepl("periods_output", directory_files_irl)])
periods_irl = read.csv(file = file_link_irl)

periods_range_irl = data.frame(Period = 1:ncol(periods_irl))
for (i in 1:ncol(periods_irl)) {
  periods_range_irl$range_years[i] = paste(min(periods_irl[, i], na.rm = T), 
                                  max(periods_irl[ , i], na.rm = T), sep = "-")
}
periods_range_irl$range_years = as.factor(periods_range_irl$range_years)

## JPN
directory_files_jpn = list.files(here("13.  bias assessment results", "jpn_invaded_clean"))
file_link_jpn = here("13.  bias assessment results", "jpn_invaded_clean", directory_files_jpn[grepl("periods_output", directory_files_jpn)])
periods_jpn = read.csv(file = file_link_jpn)

periods_range_jpn = data.frame(Period = 1:ncol(periods_jpn))
for (i in 1:ncol(periods_jpn)) {
  periods_range_jpn$range_years[i] = paste(min(periods_jpn[, i], na.rm = T), 
                                  max(periods_jpn[ , i], na.rm = T), sep = "-")
}
periods_range_jpn$range_years = as.factor(periods_range_jpn$range_years)

## MDG
directory_files_mdg = list.files(here("13.  bias assessment results", "mdg_invaded_clean"))
file_link_mdg = here("13.  bias assessment results", "mdg_invaded_clean", directory_files_mdg[grepl("periods_output", directory_files_mdg)])
periods_mdg = read.csv(file = file_link_mdg)

periods_range_mdg = data.frame(Period = 1:ncol(periods_mdg))
for (i in 1:ncol(periods_mdg)) {
  periods_range_mdg$range_years[i] = paste(min(periods_mdg[, i], na.rm = T), 
                                  max(periods_mdg[ , i], na.rm = T), sep = "-")
}
periods_range_mdg$range_years = as.factor(periods_range_mdg$range_years)

## NZL GLONAF
directory_files_nzl_glonaf = list.files(here("13.  bias assessment results", "nzl_glonaf_invaded_clean"))
file_link_nzl_glonaf = here("13.  bias assessment results", "nzl_glonaf_invaded_clean", directory_files_nzl_glonaf[grepl("periods_output", directory_files_nzl_glonaf)])
periods_nzl_glonaf = read.csv(file = file_link_nzl_glonaf)

periods_range_nzl_glonaf = data.frame(Period = 1:ncol(periods_nzl_glonaf))
for (i in 1:ncol(periods_nzl_glonaf)) {
  periods_range_nzl_glonaf$range_years[i] = paste(min(periods_nzl_glonaf[, i], na.rm = T), 
                                  max(periods_nzl_glonaf[ , i], na.rm = T), sep = "-")
}
periods_range_nzl_glonaf$range_years = as.factor(periods_range_nzl_glonaf$range_years)

## USA
directory_files_usa = list.files(here("13.  bias assessment results", "usa_invaded_clean"))
file_link_usa = here("13.  bias assessment results", "usa_invaded_clean", directory_files_usa[grepl("periods_output", directory_files_usa)])
periods_usa = read.csv(file = file_link_usa)

periods_range_usa = data.frame(Period = 1:ncol(periods_usa))
for (i in 1:ncol(periods_usa)) {
  periods_range_usa$range_years[i] = paste(min(periods_usa[, i], na.rm = T), 
                                  max(periods_usa[ , i], na.rm = T), sep = "-")
}
periods_range_usa$range_years = as.factor(periods_range_usa$range_years)

## ZAF
directory_files_zaf = list.files(here("13.  bias assessment results", "zaf_invaded_clean"))
file_link_zaf = here("13.  bias assessment results", "zaf_invaded_clean", directory_files_zaf[grepl("periods_output", directory_files_zaf)])
periods_zaf = read.csv(file = file_link_zaf)

periods_range_zaf = data.frame(Period = 1:ncol(periods_zaf))
for (i in 1:ncol(periods_zaf)) {
  periods_range_zaf$range_years[i] = paste(min(periods_zaf[, i], na.rm = T), 
                                  max(periods_zaf[ , i], na.rm = T), sep = "-")
}
periods_range_zaf$range_years = as.factor(periods_range_zaf$range_years)
```

The table below summarizes the periods covered in each of the assessed occurrence datasets

Table 1: Periods assessed 
```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
## Summary of occurrence periods
Occurrence_summary = rbind(
  c("aus_glonaf", min(periods_aus_glonaf), max(periods_aus_glonaf), ncol(periods_aus_glonaf)),
  c("gbr", min(periods_gbr), max(periods_gbr), ncol(periods_gbr)),
  c("irl", min(periods_irl), max(periods_irl), ncol(periods_irl)),
  c("jpn", min(periods_jpn), max(periods_jpn), ncol(periods_jpn)),
  c("mdg", min(periods_mdg), max(periods_mdg), ncol(periods_mdg)),
  c("nzl_glonaf", min(periods_nzl_glonaf), max(periods_nzl_glonaf), ncol(periods_nzl_glonaf)),
  c("usa", min(periods_usa), max(periods_usa), ncol(periods_usa)),
  c("zaf", min(periods_zaf), max(periods_zaf), ncol(periods_zaf))) |> 
  data.frame()

names(Occurrence_summary) = c("Range", "start", "end", "number of periods")

Occurrence_summary = Occurrence_summary |> 
  dplyr::mutate(start = as.numeric(start), 
                end = as.numeric(end),
                `number of periods` = as.numeric(`number of periods`)) |> 
  dplyr::arrange(start) 

Occurrence_summary |>
  knitr::kable()
```

Occurrence records on the invaded ranges assessed cover *3 centuries* with the earliest records taken in the *USA* and *GBR* in the *17th century* (*1601* and *1631* respectively) while *IRL*, *MDG* and *JPN* had the most recent earliest record captured in the *19th* century (*1801*, *1821* and *1861* respectively). 

We use ANOVA and the Tukey multiple comparison tests to test for variations in the levels of bias between the different *periods* (temporal bias), between *occurrence ranges*, and between the different *taxonomic groups* (*species* and *families*) which form the basis of our bias comparisons. 

# 1. Sampling intensity biases
We use the following measures to assess Sampling intensity bias:
- differences in the number of records identified between taxonomic groups
- changes in the number of records identified over time (temporal Sampling intensity bias)

## (a) Number of records identified (Sampling Intensity)
### (i) by taxonomic group **(Family)** and **(Species)**
## (b) Temporal biases in Sampling Intensity

# 2. Taxonomic Sampling biases
We use the following measures to assess taxonomic sampling bias:
- differences in the number of species identified between taxonomic groups
- changes in the number of species identified over time (temporal Sampling intensity bias)
- whether rare species are under/over-represented in the data 

## (a) Number of species identified (Taxonomic Sampling Intensity)
### (i) by taxonomic group **(Family)**
## (b) Temporal biases in Taxonomic Sampling Intensity
## (c) representation of rare species (Taxonomic Coverage Index)
## (d) Temporal changes in Taxonomic Coverage Index

# 3. Spatial biases
We use the following measures to assess taxonomic Spatial biases:
- randomness of sampling distribution in the geographic space
- temporal changes in randomness of sampling distribution in the geographic space
- temporal changes in randomness of sampling distribution in the geographic space
- spatio-temporal changes in representative portion of the spatial domain

## (a) randomness of sampling distribution in the geographic space

Uses Nearest Neighbor Index to evaluate and Indicate whether the data resemble a random distribution in the geographic space of interest for inference, and whether the extent to which the data resemble a random distribution changes over time

### (i) aus_glonaf

#### by family 

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_aus_glonaf = list.files(here("13.  bias assessment results", "aus_glonaf_invaded_clean"))

file_link_aus_glonaf = here("13.  bias assessment results", "aus_glonaf_invaded_clean", directory_files_aus_glonaf[grepl("assessSpatialBias_output", directory_files_aus_glonaf)])

file_link_aus_glonaf = file_link_aus_glonaf[!grepl("species", file_link_aus_glonaf)]
assessSpatialBias_aus_glonaf = read.csv(file = file_link_aus_glonaf)
assessSpatialBias_aus_glonaf = assessSpatialBias_aus_glonaf |> 
  merge(periods_range_aus_glonaf, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for aus_glonaf_invaded_clean occurrence data

assessSpatialBias_aus_glonaf_plot = assessSpatialBias_aus_glonaf |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by family") +
  theme(legend.position = "none")

assessSpatialBias_aus_glonaf_plot
```

#### by species 

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_aus_glonaf = list.files(here("13.  bias assessment results", "aus_glonaf_invaded_clean"))

file_link_aus_glonaf = here("13.  bias assessment results", "aus_glonaf_invaded_clean", directory_files_aus_glonaf[grepl("assessSpatialBias_output", directory_files_aus_glonaf)])

file_link_aus_glonaf = file_link_aus_glonaf[grepl("species", file_link_aus_glonaf)]
assessSpatialBias_aus_glonaf_species = read.csv(file = file_link_aus_glonaf)
assessSpatialBias_aus_glonaf_species = assessSpatialBias_aus_glonaf_species |> 
  merge(periods_range_aus_glonaf, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for aus_glonaf_invaded_clean occurrence data

assessSpatialBias_aus_glonaf_plot = assessSpatialBias_aus_glonaf_species |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by species") +
  theme(legend.position = "none")

assessSpatialBias_aus_glonaf_plot
```

### (ii) gbr

#### by family

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_gbr = list.files(here("13.  bias assessment results", "gbr_invaded_clean"))

file_link_gbr = here("13.  bias assessment results", "gbr_invaded_clean", directory_files_gbr[grepl("assessSpatialBias_output", directory_files_gbr)])

file_link_gbr = file_link_gbr[!grepl("species", file_link_gbr)]
assessSpatialBias_gbr = read.csv(file = file_link_gbr)
assessSpatialBias_gbr = assessSpatialBias_gbr |> 
  merge(periods_range_gbr, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for gbr_invaded_clean occurrence data

assessSpatialBias_gbr_plot = assessSpatialBias_gbr |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by family") +
  theme(legend.position = "none")

assessSpatialBias_gbr_plot
```

#### by species

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_gbr = list.files(here("13.  bias assessment results", "gbr_invaded_clean"))

file_link_gbr = here("13.  bias assessment results", "gbr_invaded_clean", directory_files_gbr[grepl("assessSpatialBias_output", directory_files_gbr)])

file_link_gbr = file_link_gbr[grepl("species", file_link_gbr)]
assessSpatialBias_gbr_species = read.csv(file = file_link_gbr)
assessSpatialBias_gbr_species = assessSpatialBias_gbr_species |> 
  merge(periods_range_gbr, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for gbr_invaded_clean occurrence data

assessSpatialBias_gbr_plot = assessSpatialBias_gbr_species |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by species") +
  theme(legend.position = "none")

assessSpatialBias_gbr_plot
```

### (iii) irl

#### by family

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_irl = list.files(here("13.  bias assessment results", "irl_invaded_clean"))

file_link_irl = here("13.  bias assessment results", "irl_invaded_clean", directory_files_irl[grepl("assessSpatialBias_output", directory_files_irl)])

file_link_irl = file_link_irl[!grepl("species", file_link_irl)]
assessSpatialBias_irl = read.csv(file = file_link_irl)
assessSpatialBias_irl = assessSpatialBias_irl |> 
  merge(periods_range_irl, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for irl_invaded_clean occurrence data

assessSpatialBias_irl_plot = assessSpatialBias_irl |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by family") +
  theme(legend.position = "none")

assessSpatialBias_irl_plot
```

#### by species

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_irl = list.files(here("13.  bias assessment results", "irl_invaded_clean"))

file_link_irl = here("13.  bias assessment results", "irl_invaded_clean", directory_files_irl[grepl("assessSpatialBias_output", directory_files_irl)])

file_link_irl = file_link_irl[grepl("species", file_link_irl)]
assessSpatialBias_irl_species = read.csv(file = file_link_irl)
assessSpatialBias_irl_species = assessSpatialBias_irl_species |> 
  merge(periods_range_irl, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for irl_invaded_clean occurrence data

assessSpatialBias_irl_plot = assessSpatialBias_irl_species |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by species") +
  theme(legend.position = "none")

assessSpatialBias_irl_plot
```

### (iv) jpn

#### by family

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_jpn = list.files(here("13.  bias assessment results", "jpn_invaded_clean"))

file_link_jpn = here("13.  bias assessment results", "jpn_invaded_clean", directory_files_jpn[grepl("assessSpatialBias_output", directory_files_jpn)])

file_link_jpn = file_link_jpn[!grepl("species", file_link_jpn)]
assessSpatialBias_jpn = read.csv(file = file_link_jpn)
assessSpatialBias_jpn = assessSpatialBias_jpn |> 
  merge(periods_range_jpn, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for jpn_invaded_clean occurrence data

assessSpatialBias_jpn_plot = assessSpatialBias_jpn |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by family") +
  theme(legend.position = "none")

assessSpatialBias_jpn_plot
```

#### by species

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_jpn = list.files(here("13.  bias assessment results", "jpn_invaded_clean"))

file_link_jpn = here("13.  bias assessment results", "jpn_invaded_clean", directory_files_jpn[grepl("assessSpatialBias_output", directory_files_jpn)])

file_link_jpn = file_link_jpn[grepl("species", file_link_jpn)]
assessSpatialBias_jpn_species = read.csv(file = file_link_jpn)
assessSpatialBias_jpn_species = assessSpatialBias_jpn_species |> 
  merge(periods_range_jpn, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for jpn_invaded_clean occurrence data

assessSpatialBias_jpn_plot = assessSpatialBias_jpn_species |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by species") +
  theme(legend.position = "none")

assessSpatialBias_jpn_plot
```

### (v) mdg

#### by family

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_mdg = list.files(here("13.  bias assessment results", "mdg_invaded_clean"))

file_link_mdg = here("13.  bias assessment results", "mdg_invaded_clean", directory_files_mdg[grepl("assessSpatialBias_output", directory_files_mdg)])

file_link_mdg = file_link_mdg[!grepl("species", file_link_mdg)]
assessSpatialBias_mdg = read.csv(file = file_link_mdg)
assessSpatialBias_mdg = assessSpatialBias_mdg |> 
  merge(periods_range_mdg, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for mdg_invaded_clean occurrence data

assessSpatialBias_mdg_plot = assessSpatialBias_mdg |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by family") +
  theme(legend.position = "none")

assessSpatialBias_mdg_plot
```

#### by species

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_mdg = list.files(here("13.  bias assessment results", "mdg_invaded_clean"))

file_link_mdg = here("13.  bias assessment results", "mdg_invaded_clean", directory_files_mdg[grepl("assessSpatialBias_output", directory_files_mdg)])

file_link_mdg = file_link_mdg[grepl("species", file_link_mdg)]
assessSpatialBias_mdg_species = read.csv(file = file_link_mdg)
assessSpatialBias_mdg_species = assessSpatialBias_mdg_species |> 
  merge(periods_range_mdg, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for mdg_invaded_clean occurrence data

assessSpatialBias_mdg_plot = assessSpatialBias_mdg_species |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by species") +
  theme(legend.position = "none")

assessSpatialBias_mdg_plot
```

### (vi) nzl_glonaf

#### by family

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_nzl_glonaf = list.files(here("13.  bias assessment results", "nzl_glonaf_invaded_clean"))

file_link_nzl_glonaf = here("13.  bias assessment results", "nzl_glonaf_invaded_clean", directory_files_nzl_glonaf[grepl("assessSpatialBias_output", directory_files_nzl_glonaf)])

file_link_nzl_glonaf = file_link_nzl_glonaf[!grepl("species", file_link_nzl_glonaf)]
assessSpatialBias_nzl_glonaf = read.csv(file = file_link_nzl_glonaf)
assessSpatialBias_nzl_glonaf = assessSpatialBias_nzl_glonaf |> 
  merge(periods_range_nzl_glonaf, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for nzl_glonaf_invaded_clean occurrence data

assessSpatialBias_nzl_glonaf_plot = assessSpatialBias_nzl_glonaf |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by family") +
  theme(legend.position = "none")

assessSpatialBias_nzl_glonaf_plot
```

#### by species

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_nzl_glonaf = list.files(here("13.  bias assessment results", "nzl_glonaf_invaded_clean"))

file_link_nzl_glonaf = here("13.  bias assessment results", "nzl_glonaf_invaded_clean", directory_files_nzl_glonaf[grepl("assessSpatialBias_output", directory_files_nzl_glonaf)])

file_link_nzl_glonaf = file_link_nzl_glonaf[grepl("species", file_link_nzl_glonaf)]
assessSpatialBias_nzl_glonaf_species = read.csv(file = file_link_nzl_glonaf)
assessSpatialBias_nzl_glonaf_species = assessSpatialBias_nzl_glonaf_species |> 
  merge(periods_range_nzl_glonaf, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for nzl_glonaf_invaded_clean occurrence data

assessSpatialBias_nzl_glonaf_plot = assessSpatialBias_nzl_glonaf_species |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by species") +
  theme(legend.position = "none")

assessSpatialBias_nzl_glonaf_plot
```

### (vii) usa

#### by family

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_usa = list.files(here("13.  bias assessment results", "usa_invaded_clean"))

file_link_usa = here("13.  bias assessment results", "usa_invaded_clean", directory_files_usa[grepl("assessSpatialBias_output", directory_files_usa)])

file_link_usa = file_link_usa[!grepl("species", file_link_usa)]
assessSpatialBias_usa = read.csv(file = file_link_usa)
assessSpatialBias_usa = assessSpatialBias_usa |> 
  merge(periods_range_usa, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for usa_invaded_clean occurrence data

assessSpatialBias_usa_plot = assessSpatialBias_usa |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by family") +
  theme(legend.position = "none")

assessSpatialBias_usa_plot
```

#### by species

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_usa = list.files(here("13.  bias assessment results", "usa_invaded_clean"))

file_link_usa = here("13.  bias assessment results", "usa_invaded_clean", directory_files_usa[grepl("assessSpatialBias_output", directory_files_usa)])

file_link_usa = file_link_usa[grepl("species", file_link_usa)]
assessSpatialBias_usa_species = read.csv(file = file_link_usa)
assessSpatialBias_usa_species = assessSpatialBias_usa_species |> 
  merge(periods_range_usa, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for usa_invaded_clean occurrence data

assessSpatialBias_usa_plot = assessSpatialBias_usa_species |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by species") +
  theme(legend.position = "none")

assessSpatialBias_usa_plot
```


### (viii) zaf

#### by family

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_zaf = list.files(here("13.  bias assessment results", "zaf_invaded_clean"))

file_link_zaf = here("13.  bias assessment results", "zaf_invaded_clean", directory_files_zaf[grepl("assessSpatialBias_output", directory_files_zaf)])

file_link_zaf = file_link_zaf[!grepl("species", file_link_zaf)]
assessSpatialBias_zaf = read.csv(file = file_link_zaf)
assessSpatialBias_zaf = assessSpatialBias_zaf |> 
  merge(periods_range_zaf, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for zaf_invaded_clean occurrence data

assessSpatialBias_zaf_plot = assessSpatialBias_zaf |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by family") +
  theme(legend.position = "none")

assessSpatialBias_zaf_plot
```

#### by species

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
directory_files_zaf = list.files(here("13.  bias assessment results", "zaf_invaded_clean"))

file_link_zaf = here("13.  bias assessment results", "zaf_invaded_clean", directory_files_zaf[grepl("assessSpatialBias_output", directory_files_zaf)])

file_link_zaf = file_link_zaf[grepl("species", file_link_zaf)]
assessSpatialBias_zaf_species = read.csv(file = file_link_zaf)
assessSpatialBias_zaf_species = assessSpatialBias_zaf_species |> 
  merge(periods_range_zaf, by.x = "Period", by.y = "Period") |>
  dplyr::filter(!is.na(mean)) |>
  dplyr::rename(Taxonomic_Group = identifier)

### Figure: Nearest neighbour index for zaf_invaded_clean occurrence data

assessSpatialBias_zaf_plot = assessSpatialBias_zaf_species |>
  ggplot(mapping = aes(x = Period, y = mean, color = Taxonomic_Group)) +
  geom_point( ) + geom_line() + 
  theme_bw() + labs(y = "Nearest neighbour index", 
                    title = "Randomness of the Occurrence records grouped by species") +
  theme(legend.position = "none")

assessSpatialBias_zaf_plot
```


```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
NN_index_summary = dplyr::tibble(
  
  region = c(
    "aus_glonaf", "gbr", "irl", "jpn", "mdg", "nzl_glonaf", "usa", "zaf"),
  
  `unique periods` = c(
    assessSpatialBias_aus_glonaf$Period |> unique() |> length(),
    assessSpatialBias_gbr$Period |> unique() |> length(),
    assessSpatialBias_irl$Period |> unique() |> length(),
    assessSpatialBias_jpn$Period |> unique() |> length(),
    assessSpatialBias_mdg$Period |> unique() |> length(),
    assessSpatialBias_nzl_glonaf$Period |> unique() |> length(),
    assessSpatialBias_usa$Period |> unique() |> length(),
    assessSpatialBias_zaf$Period |> unique() |> length()
  ),
  
  `unique families` = c(
    assessSpatialBias_aus_glonaf$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_gbr$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_irl$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_jpn$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_mdg$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_nzl_glonaf$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_usa$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_zaf$Taxonomic_Group |> unique() |> length()
  ),
  
  `events with index < 1` = c(
    assessSpatialBias_aus_glonaf |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_gbr |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_irl |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_jpn |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_mdg |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_nzl_glonaf |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_usa |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_zaf |> filter(mean  < 1) |> nrow()),
  
  `events with index > 1` = c(
    assessSpatialBias_aus_glonaf |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_gbr |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_irl |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_jpn |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_mdg |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_nzl_glonaf |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_usa |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_zaf |> filter(mean  > 1) |> nrow()),
  
  `min index` = c(
    assessSpatialBias_aus_glonaf$mean |> min(),
    assessSpatialBias_gbr$mean |> min(),
    assessSpatialBias_irl$mean |> min(),
    assessSpatialBias_jpn$mean |> min(),
    assessSpatialBias_mdg$mean |> min(),
    assessSpatialBias_nzl_glonaf$mean |> min(),
    assessSpatialBias_usa$mean |> min(),
    assessSpatialBias_zaf$mean |> min()),
  
  
  `max index` = c(
    assessSpatialBias_aus_glonaf$mean |> max(),
    assessSpatialBias_gbr$mean |> max(),
    assessSpatialBias_irl$mean |> max(),
    assessSpatialBias_jpn$mean |> max(),
    assessSpatialBias_mdg$mean |> max(),
    assessSpatialBias_nzl_glonaf$mean |> max(),
    assessSpatialBias_usa$mean |> max(),
    assessSpatialBias_zaf$mean |> max()))


NN_index_summary |> knitr::kable(caption = "NNI Index summary By Family")
```

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
NN_index_summary = dplyr::tibble(
  
  region = c(
    "aus_glonaf", "gbr", "irl", "jpn", "mdg", "nzl_glonaf", "usa", "zaf"),
  
  `unique periods` = c(
    assessSpatialBias_aus_glonaf_species$Period |> unique() |> length(),
    assessSpatialBias_gbr_species$Period |> unique() |> length(),
    assessSpatialBias_irl_species$Period |> unique() |> length(),
    assessSpatialBias_jpn_species$Period |> unique() |> length(),
    assessSpatialBias_mdg_species$Period |> unique() |> length(),
    assessSpatialBias_nzl_glonaf_species$Period |> unique() |> length(),
    assessSpatialBias_usa_species$Period |> unique() |> length(),
    assessSpatialBias_zaf_species$Period |> unique() |> length()
  ),
  
  `unique families` = c(
    assessSpatialBias_aus_glonaf_species$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_gbr_species$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_irl_species$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_jpn_species$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_mdg_species$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_nzl_glonaf_species$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_usa_species$Taxonomic_Group |> unique() |> length(),
    assessSpatialBias_zaf_species$Taxonomic_Group |> unique() |> length()
  ),
  
  `events with index < 1` = c(
    assessSpatialBias_aus_glonaf_species |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_gbr_species |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_irl_species |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_jpn_species |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_mdg_species |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_nzl_glonaf_species |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_usa_species |> filter(mean  < 1) |> nrow(),
    assessSpatialBias_zaf_species |> filter(mean  < 1) |> nrow()),
  
  `events with index > 1` = c(
    assessSpatialBias_aus_glonaf_species |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_gbr_species |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_irl_species |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_jpn_species |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_mdg_species |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_nzl_glonaf_species |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_usa_species |> filter(mean  > 1) |> nrow(),
    assessSpatialBias_zaf_species |> filter(mean  > 1) |> nrow()),
  
  `min index` = c(
    assessSpatialBias_aus_glonaf_species$mean |> min(),
    assessSpatialBias_gbr_species$mean |> min(),
    assessSpatialBias_irl_species$mean |> min(),
    assessSpatialBias_jpn_species$mean |> min(),
    assessSpatialBias_mdg_species$mean |> min(),
    assessSpatialBias_nzl_glonaf_species$mean |> min(),
    assessSpatialBias_usa_species$mean |> min(),
    assessSpatialBias_zaf_species$mean |> min()),
  
  
  `max index` = c(
    assessSpatialBias_aus_glonaf_species$mean |> max(),
    assessSpatialBias_gbr_species$mean |> max(),
    assessSpatialBias_irl_species$mean |> max(),
    assessSpatialBias_jpn_species$mean |> max(),
    assessSpatialBias_mdg_species$mean |> max(),
    assessSpatialBias_nzl_glonaf_species$mean |> max(),
    assessSpatialBias_usa_species$mean |> max(),
    assessSpatialBias_zaf_species$mean |> max()))


NN_index_summary |> knitr::kable(caption = "NNI Index summary By Species")
```


The Nearest neighbor index was  < 1 for all taxonomic groups across most of the periods. in all the 8 ranges. This implies that observations were more clustered than the random distribution signaling a bias towards certain sampling locations (oversampling). 
There were only a few instances where certain families had  Nearest neighbor indices > 1, i.e sampling locations were sparsely distributed relative to a normal distribution (undersampling). 

NOTE: There are more instances of undersampling than oversampling among taxonomic groups. However, there are more instances of spatial (geographic) oversampling than undersampling.

## (b) temporal changes in randomness of sampling distribution in the geographic space

### by family

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
library(broom)
library(purrr)

# aus_glonaf
spatial_index_slope_aus_glonaf = split.data.frame(assessSpatialBias_aus_glonaf, assessSpatialBias_aus_glonaf$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_aus_glonaf$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "aus_glonaf")

# gbr
spatial_index_slope_gbr = split.data.frame(assessSpatialBias_gbr, assessSpatialBias_gbr$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_gbr$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "gbr")

# irl
spatial_index_slope_irl = split.data.frame(assessSpatialBias_irl, assessSpatialBias_irl$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_irl$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "irl")

# jpn
spatial_index_slope_jpn = split.data.frame(assessSpatialBias_jpn, assessSpatialBias_jpn$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_jpn$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "jpn")

# mdg
spatial_index_slope_mdg = split.data.frame(assessSpatialBias_mdg, assessSpatialBias_mdg$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_mdg$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "mdg")

# nzl_glonaf
spatial_index_slope_nzl_glonaf = split.data.frame(assessSpatialBias_nzl_glonaf, assessSpatialBias_nzl_glonaf$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_nzl_glonaf$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "nzl_glonaf")

# usa
spatial_index_slope_usa = split.data.frame(assessSpatialBias_usa, assessSpatialBias_usa$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_usa$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "usa")

# zaf
spatial_index_slope_zaf = split.data.frame(assessSpatialBias_zaf, assessSpatialBias_zaf$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_zaf$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "zaf")

spatial_index_slope_joined = spatial_index_slope_aus_glonaf |>
  full_join(spatial_index_slope_gbr) |>
  full_join(spatial_index_slope_irl) |> 
  full_join(spatial_index_slope_jpn) |>
  full_join(spatial_index_slope_mdg) |>
  full_join(spatial_index_slope_nzl_glonaf) |>
  full_join(spatial_index_slope_usa) |>
  full_join(spatial_index_slope_zaf)

spatial_index_slope_joined |> 
  group_by(region) |> 
  summarise(families_with_increasing_index = length(which(estimate > 0)),
            families_with_decreasing_index = length(which(estimate < 0)),
            min_slope = min(estimate),
            max_slope = max(estimate)) |> 
  knitr::kable()
```


### by species

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
library(broom)
library(purrr)

# aus_glonaf
spatial_index_slope_aus_glonaf = split.data.frame(assessSpatialBias_aus_glonaf_species, assessSpatialBias_aus_glonaf_species$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_aus_glonaf$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "aus_glonaf")

# gbr
spatial_index_slope_gbr = split.data.frame(assessSpatialBias_gbr_species, assessSpatialBias_gbr_species$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_gbr$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "gbr")

# irl
spatial_index_slope_irl = split.data.frame(assessSpatialBias_irl_species, assessSpatialBias_irl_species$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_irl$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "irl")

# jpn
spatial_index_slope_jpn = split.data.frame(assessSpatialBias_jpn_species, assessSpatialBias_jpn_species$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_jpn$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "jpn")

# mdg
spatial_index_slope_mdg = split.data.frame(assessSpatialBias_mdg_species, assessSpatialBias_mdg_species$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_mdg$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "mdg")

# nzl_glonaf
spatial_index_slope_nzl_glonaf = split.data.frame(assessSpatialBias_nzl_glonaf_species, assessSpatialBias_nzl_glonaf_species$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_nzl_glonaf$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "nzl_glonaf")

# usa
spatial_index_slope_usa = split.data.frame(assessSpatialBias_usa_species, assessSpatialBias_usa_species$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_usa$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "usa")

# zaf
spatial_index_slope_zaf = split.data.frame(assessSpatialBias_zaf_species, assessSpatialBias_zaf_species$Taxonomic_Group) |>
  purrr::map(~lm(mean ~ 1 + Period, data = .x))|>
  purrr::map_df(tidy) |>
  dplyr::filter(term == 'Period') |>
  dplyr::mutate(term = sort(assessSpatialBias_zaf$Taxonomic_Group |> unique())) |>
  dplyr::filter(p.value < 0.05 | p.value == 0.05) |>
  dplyr::rename(Taxonomic_group = term) |> dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(region = "zaf")

spatial_index_slope_joined = spatial_index_slope_aus_glonaf |>
  full_join(spatial_index_slope_gbr) |>
  full_join(spatial_index_slope_irl) |> 
  full_join(spatial_index_slope_jpn) |>
  full_join(spatial_index_slope_mdg) |>
  full_join(spatial_index_slope_nzl_glonaf) |>
  full_join(spatial_index_slope_usa) |>
  full_join(spatial_index_slope_zaf)

spatial_index_slope_joined |> 
  group_by(region) |> 
  summarise(families_with_increasing_index = length(which(estimate > 0)),
            families_with_decreasing_index = length(which(estimate < 0)),
            min_slope = min(estimate),
            max_slope = max(estimate)) |> 
  knitr::kable()
```


## (c) spatio-temporal changes in representative portion of the spatial domain

Indicates the extent to which a representative portion of the spatial domain of interest has been sampled over time. Output is a single map for each taxonomic group showing the number of time periods in which each grid cell has been sampled.

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
library(knitr)
library(ggplot2)
library(gridExtra)
library(raster)

links_to_images = data.frame()

for(region in tables) {
  species_folder = here("13.  bias assessment results", region, "spatial_cov_species")
  family_folder = here("13.  bias assessment results", region, "spatial_cov_family")
  
  paste(region, ": number of time periods each grid cell was sampled") |> print()
  
  "By Family" |> print()
  
  links_to_images = rbind(
    links_to_images,
    list(region = region,
           group = "family",
           links = here(family_folder)) |> as.data.frame())
  
  "By Species" |> print()
  
  links_to_images = rbind(
    links_to_images,
    list(region = region,
           group = "species",
           links = here(species_folder)) |> as.data.frame())
}

links_to_images |> knitr::kable()
```


# 4. Environment Bias

```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
## aus_glonaf
env_bias_summary = function(db_table) {
  directory_files = list.files(here("13.  bias assessment results", paste0(db_table, "_invaded_clean")))
  file_links = here("13.  bias assessment results", paste0(db_table, "_invaded_clean"), directory_files_aus_glonaf[grepl("assessEnvBias_output", directory_files_aus_glonaf)])
  
  # Plot data
  
  ### by family
  file_link_family = file_link_aus_glonaf[!grepl("species", file_link_aus_glonaf)] |> readr::read_csv()
  if(nrow(file_link_family) > 0) {
    # Assuming 'zaf' is your data frame
family_plot = ggplot(file_link_family, aes(x = scores.PC1, y = scores.PC2, color = Period, group = Period)) +
  stat_ellipse(type = "norm", geom = "polygon", alpha = 0.25) +
  labs(title = "Scatter Plot with Ellipses by family",
       x = "PC1",
       y = "PC2")
  }
    
   ### by species
  file_link_species = file_link_aus_glonaf[grepl("species", file_link_aus_glonaf)] |> readr::read_csv()
  if(nrow(file_link_species) > 0) {
    # Assuming 'zaf' is your data frame
species_plot = ggplot(file_link_species, aes(x = scores.PC1, y = scores.PC2, color = Period, group = Period)) +
  stat_ellipse(type = "norm", geom = "polygon", alpha = 0.05) +
  labs(title = "Scatter Plot with Ellipses by Species",
       x = "PC1",
       y = "PC2")
  }
  
  return(list(
    if(exists("family_plot")) {
      family_plot = family_plot
    },
    
    if(exists("species_plot")) {
      species_plot = species_plot
    }
  ))
}
```


```{r, echo =FALSE, cache=TRUE, results='asis', message=FALSE, warning=FALSE}
env_bias_summary(db_table = "zaf")
```





